---
title: "Example for Version 2 of Paper"
output: html_notebook
---

I would like to create a new example that is realistic but still simple enought that it can be verified manually.

Begin by clearing the workspace:
```{r clear-workspace, echo=FALSE}
rm(list = ls())
```

Next, load the packages needed for this:
```{r load-packages, echo=FALSE, include=FALSE}
library(tidyverse)
library(rgdal)
library(raster)
library(sf)
library(dplyr)
library(gdistance)
library(kableExtra)
```

Load geography:
```{r retrieve-geography-files, echo=FALSE, include=FALSE}
load("Simulated Example.RData")
```

Create simple features objects for population centers and clinics:
```{r}
# Clinics
sim_clinics <- st_sf(Clinics_data, geometry_clinics, stringsAsFactors = TRUE)

# Population centers
sim_population <- st_sf(Population_data, geometry_population, stringsAsFactors = TRUE)
```

Calculate the distance between these points as a measure of cost:
```{r}
distance_matrix <- round(pointDistance(st_coordinates(sim_population), 
                                 st_coordinates(sim_clinics), 
                                 type='Euclidean', lonlat = FALSE))
```

Create buffers around the clinics using some threshold value, say the 4th quintile of distances in the system:
```{r}
# Clinic
clinic_buffers_1 <- st_buffer(sim_clinics, quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[5])

clinic_buffers_2 <- st_buffer(sim_clinics, quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[3])

clinic_buffers_3 <- st_buffer(sim_clinics, quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[1])

# Population centers
population_buffers <- st_buffer(sim_population, quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[4])
```

Plot the first stage of the 2SFCA algorithm (i.e., clinic catchments):
```{r}
ggplot() +
  geom_sf(data = clinic_buffers_1, aes(fill = id), alpha = 0.3) +
  geom_sf(data = clinic_buffers_2, aes(fill = id), alpha = 0.3) +
  geom_sf(data = clinic_buffers_3, aes(fill = id), alpha = 0.3) +
  geom_sf(data = sim_population, aes(size = Population, color = as.factor(Population))) +
  scale_size(guide = FALSE) +
  guides(fill = guide_legend(title = "Clinic"), color=guide_legend(title="Population")) +
  geom_sf_text(data = sim_population, aes(label = id), vjust = -1.5, size = 3) +
  geom_sf(data = sim_clinics, shape = 17, color = "red", size = 3)+
  theme_minimal()
```

Notice that some population centers are inside the catchment areas of more than one clinic. For instance, Population Center 4 is in the catchment areas of Clinics 1 and 2. To see how this impacts the calculations, define an impedance matrix using the same criteria as for the buffers:
```{r}
# Quantiles according to distance
quantiles_d <- quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))

impedance_binary <- 1 * (distance_matrix <= quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[5])

impedance_stepwise <- 0.945 * (distance_matrix <= quantiles_d[1]) + 
  0.600 * (distance_matrix > quantiles_d[1] & distance_matrix <= quantiles_d[3]) +
  0.242 * (distance_matrix > quantiles_d[3] & distance_matrix <= quantiles_d[5]) +
  0 * (distance_matrix > quantiles_d[5])
  
kable(data.frame(Center = Population_data$id, impedance_binary, impedance_stepwise), 
      "html", 
      col.names = c("Population Center",
                           "Clinic 1",
                           "Clinic 2",
                           "Clinic 3",
                           "Clinic 1",
                           "Clinic 2",
                           "Clinic 3")) %>%
  add_header_above(c(" " = 1, "Binary Impedance" = 3, "Stepwise Impedance" = 3))
```

The demand can be calculated as:
```{r}
demand_binary <- (t(impedance_binary) %*% (sim_population$Population))
demand_stepwise <- (t(impedance_stepwise) %*% (sim_population$Population))

kable(data.frame(Clinic = c("Clinic 1", "Clinic 2", "Clinic 3"), 
                 demand_binary = demand_binary, 
                 demand_stepwise = demand_stepwise), 
      digits = 3,
      "html",
      col.names = c("Clinic",
                    "Demand (Binary)",
                    "Demand (Stepwise)"))
```

It can be seen that the total population in this system is `r sum(sim_population$Population)` and yet, the assumed level of demand is `r sum(demand_binary)`. This is because the algorithm assumes that every person can potentially attend more than one clinic. This tends to inflate the demand for the service.

Calculate row and column adjusted-impedance:
```{r}
#Row standardize the impedance matrix to ensure that the sum of the population over the clinics preserves the population totals
row_sums_binary <- rowSums(impedance_binary) #sum weights over rows
row_sums_binary[row_sums_binary == 0] <- 1 #change to one for those rows where the total is zero to avoid divisions by zero
impedance_binary_row_adjusted <- sweep(impedance_binary, 1, row_sums_binary, FUN = "/") #row standardize

#Column standardize the impedance matrix to ensure that the sum of the level of service preserves the total supply
col_sums_binary <- colSums(impedance_binary) #sum weights over rows
col_sums_binary[col_sums_binary == 0] <- 1 #change to one for those rows where the total is zero to avoid divisions by zero
impedance_binary_col_adjusted <- sweep(impedance_binary, 2, col_sums_binary, FUN = "/") #column standardize

kable(data.frame(Center = Population_data$id, impedance_binary_row_adjusted, impedance_binary_col_adjusted), 
      "html", 
      digits = 3,
      col.names = c("Population Center",
                           "Clinic 1",
                           "Clinic 2",
                           "Clinic 3",
                           "Clinic 1",
                           "Clinic 2",
                           "Clinic 3")) %>%
  add_header_above(c(" " = 1, "Row Standardized" = 3, "Column Standardized" = 3))
```

The 3SFCA of Wan et al. (2012) uses a selection weight to adjust the population. The selection weights are essentially a row standardized matrix of weights. They can be based on the cost matrix, as follows (using a negative exponential function to calculate the weights): 
```{r}
selection_weights <- exp(-0.0000001 * distance_matrix^2)
```

Moreover, the selection weights apply only to centers that are within the catchment zone, as per the impedance function:
```{r}
selection_weights <- selection_weights * impedance_binary
```

Finally, the selection weights are standardized so that the sum of the rows is one:

$$
G_{ij}=\frac{T(d_{ij})}{\sum_jT(d_{ij})}
$$

```{r}
row_sums_binary <- rowSums(selection_weights) #sum weights over rows
row_sums_binary[row_sums_binary == 0] <- 1 #change to one for those rows where the total is zero to avoid divisions by zero
selection_weights <- sweep(selection_weights, 1, row_sums_binary, FUN = "/") #row standardize

kable(data.frame(Center = Population_data$id, selection_weights), 
      "html", 
      digits = 3,
      col.names = c("Population Center",
                           "Clinic 1",
                           "Clinic 2",
                           "Clinic 3"))
```

The level of service is calculated as follows:
```{r}
demand_selection_binary <- (t(selection_weights * impedance_binary) %*% (sim_population$Population))
demand_selection_stepwise <- (t(selection_weights * impedance_stepwise) %*% (sim_population$Population))

kable(data.frame(Clinic = c("Clinic 1", "Clinic 2", "Clinic 3"), 
                 demand_selection_binary = demand_selection_binary,
                 demand_selection_stepwise = demand_selection_stepwise),
      digits = 3,
      "html",
      col.names = c("Clinic",
                    "Demand (Binary)",
                    "Demand (Stepwise)"))
```

